---
title: "STAT 240 Final Project"
output: html_document
---

## **Team Members**
Matthew Chiang, Siyu Huang, Vishnu Yaragadda

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
source("./scripts/viridis.R")
source("./scripts/ggprob.R")
```

```{r cache=TRUE, cache.path="cache/", include=FALSE}
# get all dates
file_list <- list.files(path="./data")

# load all data
allData <- data.frame()
for(i in 1:length(file_list)){
  
  # get file names
  league <- paste("./data/", file_list[i], "/LeagueTable.csv", sep="")
  standard <- paste("./data/", file_list[i], "/SquadStandardStats.csv", sep="")
  goalkeep <- paste("./data/", file_list[i], "/SquadGoalkeeping.csv", sep="")
  
  # load in each data table and merge it together
  LeagueTable <- read.csv(league)
  if( file.exists(standard) ){
    SquadStandardStats <- read.csv(standard)
    LeagueTable <- full_join(LeagueTable, SquadStandardStats, by="Squad")
  }
  if( file.exists(goalkeep) ){
    SquadGoalkeeping <- read.csv(goalkeep)
    LeagueTable <- full_join(LeagueTable, SquadGoalkeeping, by="Squad")
  }
  
  
  # add year column to the data
  LeagueTable <- select(LeagueTable, -contains(".y")) %>%
    mutate(year = substr(file_list[i], 6, 9))
  
  
  LeagueTable <- LeagueTable %>% 
    top_n(LeagueTable, 20)
  # combine the data into one big table
  if( i != 1){
    LeagueTable[setdiff(names(allData), names(LeagueTable))] <- NA
    allData[setdiff(names(LeagueTable), names(allData))] <- NA
  }
  allData <- rbind(allData, LeagueTable)
}
allData <- allData %>%
  select(Rk, Squad, W.x, D.x, L.x, GF, GA.x, Top.Team.Scorer, Poss, Saves,year)
```

## **Introduction**

"People cry when their team lose, and they cry even more when their team wins." Our team decided to conduct our analysis on the La Liga Football League as it is one of the most influential soccer leagues in the world. The league also has the most exciting rivalry in history: Real Madrid vs Barcelona. Our team members are also fans of this soccer league and we decided to conduct research on the La Liga data. Throughout history, only nine different teams have been crowned champions, and we are curious about what makes those teams so special. More specifically, what are some of their common characteristics that result in them always performing so well.

We theorize that three reasons why those teams perform so well is that the total amount of goals per season of said top teams keeps increasing in a consistent or linear manner, the top scorer for the teams that perform well have a higher contribution to the total goals when compared to the rest of the league, and that the possession rate for those top teams are also much higher.

## **Background**

> Raw Data:

We found data for the La Liga Soccer League. Each season from 1988-2020, there are multiple sets of data ranging from team rankings, top scorers, standard squad statistics, team goalkeeping, team shooting, to statistics on individual players. We selected three sets of data:

* League Table: includes team rankings, total wins, loses, goals for, goals against, etc.
* Squad Standard Stats: includes general performance statistics of each team.
* Squad Goalkeeping: includes performance statistics for the goalkeeping of each team.

> Final Data: 

We had to collect the above data for the seasons between 1988-2020. We combined all of the data into one large dataframe called allData by the team name and created a year column to keep track of which teams data corresponds to which year. Then, we trimmed most of the variables in that large data set away and only selected the ones that were used in the analysis. 

**allData** contains the variables described below:

Variable| Description
----------------|------------
year|Season year 
Rk|Rank of each team for that season
Squad|Team Name
W.x|Number of wins for that team during season
L.x|Number of losses for that team during season
D.x|Number of draws for that team during season
GF|Total number of goals the team scored during season
GA.x|Total number of goals the team conceded during season
Top.Team.Scorer|Name of team's top scorer during season
Poss|Average Rate of Possession of the ball during season
Saves|Total number of saves the team had during season

> Data Citation:

"La Liga Seasons", Real Federación Española de Fútbol, 7th April 2020, https://fbref.com/en/comps/12/history/La-Liga-Seasons 
 
> Data Caveats:
 
- To perform our analysis we considered all teams data, not just the teams that are currently in the league.
- The data set we found does not have data for Poss or Saves before the year 2000. In parts of analysis that use those metrics, we used the data for the top 3 teams of each season instead of just the champion team. This allowed us to have a more accurate sample size.


## **Analysis**

> Question 1: 
First, we wanted to test our claim that we can make a linear regression model for the amount of total goals of the top 3 teams of the current season. 

First, we look at the number of Goals of the top 3 teams for each season over the years. 

```{r, echo = FALSE}
b <- allData %>%
  filter(allData$Saves != "N/A") %>%
  filter(Rk <= 3) %>%
  mutate(year = as.numeric(year)) %>%
  select(GF, year)
b %>%
  ggplot(aes(x = year, y = GF)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ylab("Goals") +
  ggtitle("Number of Goals of Champion")
lm1 <- lm(GF~year, b)
library(modelr)
b <- b %>%
  add_residuals(lm1) %>%
  add_predictions(lm1)
ggplot(b, aes(x=year, y =resid)) +
  geom_point() +
  xlab("Years") +
  ylab("Residuals") +
  ggtitle("Residuals for Goals~Year") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")
```

Just from looking at the graph, we can recognize that it looks it has a positive linear relationship. Then, we also look at the residuals so we can make sure those residuals follow an approximate normal distribution and have no strong pattern. Since, this is the case we can now state our linear regression model. 

$$
Goals = 1.1667*year - 2265.2698
$$
Now, we state our 95% confidence interval for the slope of the linear regression. 

```{r,include = FALSE}
confint(lm1)
upper <- -697.106 + (1.947 * 2021)
lower <- -3833.434 + (0.386 * 2021)
```

So, according to our calculations, our 95% confidence interval is between 0.3864 and 1.9468. After conduction some research we found that because 0 does not exist within our confidence interval, this shows that there exists significant evidence of a linear relationship.

> Analysis: To conclude, there is no significant evidence of a linear relationship and the statement in our thesis is proven to be false.


> Next, we wanted to look at the second part of our thesis about how a top team has a higher proportion of goals from their top scorer compared to the rest of the league.

For each champion team of each season, they have a top scorer that scores more goals than other players in the team and we believe that this consists of a sizable proportion of total goals of the team compared to other teams. 
```{R echo=FALSE}
top_scorer <- allData %>% 
  mutate(top_scorer_goals = as.numeric(str_trim(str_sub(Top.Team.Scorer, -2, -1)))) %>% 
  # mutate(t2 = str_extract(Top.Team.Scorer, "[\d+][?!.*\d]$")) %>% 
  mutate(prop = top_scorer_goals / GF) %>% 
  mutate(year = as.numeric(year)) %>% 
  select(Squad, top_scorer_goals, GF, prop, year)
# ggplot(top_scorer, aes(x=year, y=prop)) +
#   geom_point() +
#   geom_smooth(method = "lm", se = FALSE, color = "hotpink") +
#   ggtitle("The proportion of goals by top scorer in each seasons") +
#   labs(x="Season", y="The proportion of goals by top scorer")
```

```{R echo=FALSE}
top_scorer2 <- allData %>%
  mutate(Rk = case_when(Rk == 1 ~ "Champion", 
                        TRUE ~ "Not Champion",)) %>%
  mutate(top_scorer_goals = as.numeric(str_trim(str_sub(Top.Team.Scorer, -2, -1)))) %>% 
  mutate(prop = top_scorer_goals / GF) %>% 
  mutate(year = as.numeric(year)) %>% 
  select(Rk, top_scorer_goals, GF, prop, year)

not_champion_top_scorer <- top_scorer2 %>% 
  filter(Rk == "Not Champion") %>% 
  group_by(year) %>% 
  summarise(not_champion_prop = mean(prop))
champion_top_scorer <- top_scorer2 %>% 
  filter(Rk == "Champion") %>% 
  group_by(year) %>% 
  summarise(champion_prop = mean(prop)) %>% 
  inner_join(not_champion_top_scorer)
```

```{R echo=FALSE}
champion_top_scorer2 = gather(champion_top_scorer, event, total, champion_prop:not_champion_prop) %>% 
  mutate(`Season Champion` = event)
# champion_top_scorer2
ggplot(champion_top_scorer2, aes(x=`Season Champion`, y=total, fill=`Season Champion`)) +
  geom_boxplot() +
  ggtitle("Comparison between champions and other teams", subtitle = ":the proportion of goals by top scorer in each seasons") +
  labs(x="Champions or not", y="The proportion of goals by top scorer", color = "Season Champion") 
```

From the observation of the graph, we assume that the proportion of top scorer's goals in champions teams tend to be higher than those of non-champions teams. To verify our guess, we can perform a hypothesis test.
$$
H_0: p_0 = p_1
$$
$$
H_a: p_0 > p_1
$$
where $\mu_0$ is the proportion of top scorer's goals over total goals for champions teams, and $\mu_1$ is the top scorer's goals over total goals for non-champions teams.
To carry out this test, we will consider a two-sample z-test with the following test statistic:
$$
T = \frac{p_0-p_1}{\sqrt{\frac{s_0^2}{n_0} + \frac{s_1^2}{n_1}}}
$$

```{R include = FALSE}
# lm_s = lm()
# t.test(champion_top_scorer$champion_prop, champion_top_scorer$not_champion_prop, alternative = "greater")

# top_scorer2 %>% 
#   group_by(Rk) %>% 
#   summarise(sum_top_scorer = sum(top_scorer_goals),
#             sum_GF = sum(GF))
prop.test(c(858, 8118), c(2794, 29612))
```

After performing the test we get the following results:

- $T = 13.67$
- ${DegreesOfFreedom} = 1$
- $PValue = 0.0002179$

We perform the hypothesis test where the null hypothesis is that the proportions are equal.  After doing the z-test, we found that the p-value is 0.0002179 and it is statistically significant so that we can reject the null hypothesis. Therefore, The proportion of top scorer's goals of all goals in champions teams tend to be higher than those of non-champions teams which supports our thesis.

<!-- > How about the trend of this proportion we just talked about? -->

```{R echo=FALSE}
# ggplot(top_scorer2, aes(x=year, y=prop, color=as.factor(Rk))) +
#   geom_point(alpha = 0.7) +
#   geom_smooth(method="lm", se = FALSE) +
#   scale_color_manual(values = c("darkgoldenrod", "blue")) +
#   ggtitle("The proportion of goals by top scorer in each seasons") +
#   labs(x="Season", y="The proportion of goals by top scorer", color = "Season Champion")
```

> Now, our last part of the thesis states that the possession rate of a top team is higher than the rest of the league. 

It's reasonable to assume that a winning team would have a higher possession rate as that would give the team more chances to score. So, we wanted to test that:

```{r, echo = FALSE}
poss <- allData %>%
  filter(Poss != "NA") %>%
  mutate(Rk = case_when(Rk <= 3 ~ "Top Teams", 
                        TRUE ~ "Rest of League",)) %>%
  group_by(Rk) %>%
  ggplot(aes(x = Poss, y = GF, color=as.factor(Rk))) +
  geom_point() +
  scale_color_manual(values = c("#e05a5e", "#58e0d5")) +
  labs(x="Possession Rate", y="Goals Scored", color = "La Liga Teams") +
  ggtitle("Possession Rate vs Goals")

poss
```

The graph above supports our initial claim as the Top Teams have a much higher possession rate and thus end up with more goals scored. Now, we want to conduct a hypothesis test to reaffirm our claim that teams with high possession rates tend to do better. 

We can frame this as a hypothesis test described as follows:
$$
H_0: \mu_0 = \mu_1
$$
where $\mu_0$ is the mean possession rate for the rest of the league, and $\mu_1$ is the mean possession rate for the top teams.

$$
H_a: \mu_0 < \mu_1
$$

To carry out this test, we will consider a two-sample test with the following test statistic:
$$
T = \frac{(\mu_0-\mu_1)}{\sqrt{\frac{s_0^2}{n_0} + \frac{s_1^2}{n_1}}}
$$
where under the null $\mu_0 - \mu_1 = 0$, $n_i$ is the sample size for group $i=0,1$, and $s_i$ is the standard deviation of group $i$.

Since we are estimating the standard error, we will be using a t-distribution rather than a normal distribution.

```{r, include=FALSE}
poss0 <- allData %>%
  filter(Poss != "NA") %>%
  filter(Rk > 3) %>%
  pull(Poss)

poss1 <- allData %>%
  filter(Poss != "NA") %>%
  filter(Rk <= 3) %>%
  pull(Poss)

mean0 <- mean(poss0)
sd0 <- sd(poss0)
n0 <- length(poss0)

mean1 <- mean(poss1)
sd1 <- sd(poss1)
n1 <- length(poss1)

## Assuming the null is true
se <- sqrt(sd0^2/n0+sd1^2/n1)
test_stat_poss <- (mean0-mean1)/se
```
```{r echo=FALSE}
## Get degrees of freedom of t-distribution from t.test()
t.test(poss0, poss1, alternative ="less")

dof <- 19.6

pt(test_stat_poss, dof) # p-value = P(T_dof <= -9.76586)

```

After performing our t-test we get the following results:

- $T = -4.8765$
- ${DegreesOfFreedom} = 19.6$
- $PValue = 4.82*10^{-5}$

Using a t-distribution with 19.6 degrees of freedom, we calculate a p-value on the order of $10^{-5}$ suggesting that we can reject the null hypothesis. This rejections means we have concluded that the mean possession rate for the top teams is greater than the mean possession rate for the rest of the league.

## Discussion
In the world of soccer, everything is unpredictable and there are a lot more factors that could affect the trend of a season. For example, transfers could bring new players that strengthen the team and the injuries of key players could hinder a team's success. It is even more difficult to predict the winner of a future season, but we can definitively say a few things about top performing teams.

From our analysis of the La Liga data, we are able to say that the top teams will most likely have a top scorer with a higher proportion of total goals when compared to the average top scorer of the other teams. This makes sense if you consider that some of the top teams in the league have players that are some of the best to ever play the game. For example, Lionel Messi played 16 years for Barcelona and he is the top scorer in the team's history. Likewise Cristiano Ronaldo played 9 years with Real Madrid. Lastly we are also able to confirm that the top teams of the league will most likely have a higher possession rate than the rest of the league. This makes sense as the longer a team holds the ball, the more chances it has to score and it also reduces the chances of the other team to score.

One shortcoming of our project was that we had very limited amounts of data. Although we had league data from 1988 through 2020, certain metrics were not logged before 2000, so for parts of our analysis where we used data that was only available from 2000 and later, we had to account this by analyzing the top 3 teams instead of just the champion team. Consequently, our results may not be as accurate. Another shortcoming is that we tended to focus on only three aspects of a soccer team, namely goals, possession and top scorer . But, in reality, there might be a lot more variables that come into play. For example, we could consider each team's overall offensive and defensive ratings to see how that has affected the winners of each season throughout history. This question could be done in the future to expand our analysis.

To conclude, through our data analysis of the La Liga soccer league, we are able to confirm that the top performing teams have a top scorer who is more influential than his counterparts on other teams as we performed a t-test which gave us a p-value of 0.012. Then, we can also say that the possession rate of a top team will be, on average, higher than other teams in the league as we got a p-value to the order of $10^{-5}$ when performing that t-test. 
