---
title: "STAT 240 Final Project"
output: html_document
---

## **Team Members**
Matthew Chiang, Siyu Huang, Vishnu Yaragadda

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
library(modelr)
library(gtools)
source("./scripts/viridis.R")
source("./scripts/ggprob.R")
```

```{r cache=TRUE, cache.path="cache/", include=FALSE}
# get all dates
file_list <- list.files(path="./data")

# load all data
allData <- data.frame()
for(i in 1:length(file_list)){
  
  # get file names
  league <- paste("./data/", file_list[i], "/LeagueTable.csv", sep="")
  standard <- paste("./data/", file_list[i], "/SquadStandardStats.csv", sep="")
  goalkeep <- paste("./data/", file_list[i], "/SquadGoalkeeping.csv", sep="")
  
  # load in each data table and merge it together
  LeagueTable <- read.csv(league)
  if( file.exists(standard) ){
    SquadStandardStats <- read.csv(standard)
    LeagueTable <- full_join(LeagueTable, SquadStandardStats, by="Squad")
  }
  if( file.exists(goalkeep) ){
    SquadGoalkeeping <- read.csv(goalkeep)
    LeagueTable <- full_join(LeagueTable, SquadGoalkeeping, by="Squad")
  }
  
  
  # add year column to the data
  LeagueTable <- select(LeagueTable, -contains(".y")) %>%
    mutate(year = substr(file_list[i], 6, 9))
  
  LeagueTable <- LeagueTable %>% 
    top_n(LeagueTable, 20)
  # combine the data into one big table
  if( i != 1){
    LeagueTable[setdiff(names(allData), names(LeagueTable))] <- NA
    allData[setdiff(names(LeagueTable), names(allData))] <- NA
  }
  
  allData <- rbind(allData, LeagueTable)
}
allData
```

## **Introduction**

Our team decided to conduct our analysis on the La Liga Football League as we all enjoy watching soccer and thought it would be interesting to try and make some sort of prediction on the winner of the current season. We decided that it would be best to analyze past seasons to predict the characteristics of the winner of the current season. To begin out approach we will look historicaly into the league to discover the probability that they win the current season. Then we will leverage a 95% confidence interval on a normal distribution of points and wins for the winners of the league. Lastly we will perform a linear regression on the winners of goals and saves vs years to exterpolate an estimation of how many saves and goals the 2020-2021 La Liga champion will have.

## **Background**

> Raw Data:

We found data for the La Liga Soccer League. Each season from 1988-2020, there are multiple sets of data ranging from team rankings, top scorers, standard squad statistics, team goalkeeping, team shooting, to statistics on individual players. We selected three sets of data:

* League Table: includes team rankings, total wins, loses, goals for, goals against, etc.
* Squad Standard Stats: includes general performance statistics of each team.
* Squad Goalkeeping: includes performance statistics for the goalkeeping of each team.

> Final Data: 

We had to collect the above data for the seasons between 1988-2020. We combined all of the data into one large dataframe by the team name and created a year column to keep track of which teams data cooresponds to which year.

> Data Citation:

"La Liga Seasons", Real Federación Española de Fútbol, 7th April 2020, https://fbref.com/en/comps/12/history/La-Liga-Seasons 
 
> Focus:
 
The overall aim of our analysis on this data will be to determine the characteristics of the winning team in the 2020-2021 La Liga Soccer League.


## **Analysis**


> Question 1: What teams have won in the past?

First, we want to narrow our analysis by looking at the history of the La Liga Soccer League. To do this we will calculate the percentage of championship wins each team has.

```{r echo=FALSE}
championNames <- allData %>% 
  filter(Rk == 1) %>%
  group_by(Squad, Rk) %>%
  summarize(
    wins = n()
  ) %>%
  ungroup() %>%
  mutate(pr = wins/sum(wins)) %>%
  arrange(desc(Rk), desc(wins))

championNames$Squad <- reorder(championNames$Squad, -championNames$pr)

ggplot(championNames, aes(x=Squad, y=pr, fill=factor(Rk))) +
  geom_col() +
  scale_fill_manual(values = c("#e05a5e", "#58e0d5")) +
  xlab("Team") +
  ylab("Probability of Winning League") +
  ggtitle("Percentage of Championship Wins", subtitle = "by Team") +
  theme(legend.position = "none")
```

From this graph it is clear that `r championNames$Squad[1]` and `r championNames$Squad[2]` have a overwhelming win rate compared to the other teams in the league, historically winning `r (championNames$pr[1] + championNames$pr[2])*100`% of the time. With this realization, specify our analysis more narrowly to check characteristics for when the champion is `r championNames$Squad[1]`, `r championNames$Squad[2]`, or other winners.

> Question 2: 

Second, we want to analyze the most direct factor that decides La Liga's winner of each season, that is the points each team get after 38 matches. In this league system, teams receive three points for a win, one point for a draw, and no points for a loss. Teams are ranked by total points, with the highest-ranked club at the end of the season crowned champion. From `r min(allData$year)` to `r max(allData$year)`, each winner's final point are gathered and we do the confidence interval on distribution to estimate the number of points that a champion should get. 

```{R echo=FALSE}
champions <- allData %>%
  filter(Rk == 1)

champions1 <- champions%>%
  summarize(
    mean_pt = mean(Pts),
    z_pt = 1.960, #this is the 95% confidence level value
    s_pt = sd(Pts),
    n_pt = n(),
    CI_pt = mean_pt + c(-1,1) * z_pt * s_pt / sqrt(n_pt)
  )

ggplot(champions1)+
  geom_norm_density(mu=mean_pt,sigma=s_pt,a=NULL,b=NULL, color="#000000") +
  geom_norm_fill(mu=mean_pt,sigma=s_pt,a=NULL,b=NULL,fill="#000000", alpha = 0.1) +
  geom_vline(xintercept = CI_pt[1], linetype = 2, color="#888888") +
  geom_vline(xintercept = CI_pt[2], linetype = 2, color="#888888") +
  geom_text(aes(CI_pt[1],-0.002,label=round(CI_pt[1]))) +
  geom_text(aes(CI_pt[2],-0.002,label=round(CI_pt[2]))) +
  geom_hline(yintercept=0) +
  ylab('Density') +
  xlab('Points In A Season')
```

```{r echo=FALSE}
champions2 <- champions %>%
  mutate(Squad =
    case_when(
      Squad == championNames$Squad[1] ~ Squad,
      Squad == championNames$Squad[2] ~ Squad,
      TRUE ~ "Other"
    )
  ) %>%
  group_by(Squad)%>%
  summarize(
    mean_pt = mean(Pts),
    z_pt = 1.960, #this is the 95% confidence level value
    s_pt = sd(Pts),
    n_pt = n(),
    lower = mean_pt - z_pt * s_pt / sqrt(n_pt),
    upper = mean_pt + z_pt * s_pt / sqrt(n_pt)
  )

ggplot(champions2, color="#58e0d5", fill="#58e0d5")+
    geom_norm_density(mu=mean_pt,sigma=s_pt,a=NULL,b=NULL, color="#58e0d5") +
    geom_norm_fill(mu=mean_pt,sigma=s_pt,a=NULL,b=NULL,fill="#58e0d5", alpha = 0.1) +
    geom_vline(xintercept = CI_pt[1], linetype = 2, color="#888888") +
    geom_vline(xintercept = CI_pt[2], linetype = 2, color="#888888") +
    geom_text(aes(lower,-0.002,label=round(lower))) +
    geom_text(aes(upper,-0.002,label=round(upper))) +
    facet_wrap(~ Squad) +
    geom_hline(yintercept=0) +
    ylab('Density') +
    xlab('Points In A Season')
```
From the distribution graph, we estimate that the winning point of 95% confidence interval should be between approximately 74 to 84.

```{R echo=FALSE}
mean_wins = mean(champions$W.x)
z_wins = 1.960 #this is the 95% confidence level value
s_wins = sd(champions$W.x)
n_wins = nrow(champions)
CI_wins = mean_wins + c(-1,1) * z_wins * s_wins / sqrt(n_wins)
gnorm(mean_wins, s_wins, color = 'cyan') +
  geom_vline(xintercept = CI_wins[1], color = 'hotpink', linetype = 2) +
  geom_vline(xintercept = CI_wins[2], color = 'hotpink', linetype = 2) +
  geom_text(aes(CI_wins[1],-0.002,label=round(CI_wins[1]))) +
  geom_text(aes(CI_wins[2],-0.002,label=round(CI_wins[2]))) +
  ggtitle('Normal distributions of champions wins') +
  xlab('Number of Wins')
```
From the distribution graph, we estimate that the number of wins of 95% confidence interval should be between approximately 25 to 27.

> Analysis: To conclude, we will have 95% confidence to say that the champion of the current season of La Liga will win 25 to 27 matches and get 74 to 84 points.

> Question 2:

Now, we wanted to better understand the trends of the goals and saves of the champion of the LaLiga over the years.


```{r echo=FALSE}
a <- allData %>%
  filter(allData$Saves != "N/A") %>%
  filter(Rk == "1") %>%
  mutate(year = as.numeric(year)) %>%
  select(Saves, year)

a %>%
  ggplot(aes(x = year, y = Saves)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

lm0 <- lm(Saves~year, a)

a <- a %>%
  add_residuals(lm0) %>%
  add_predictions(lm0)

ggplot(a, aes(x=year, y =resid)) +
  geom_point() +
  xlab("x") +
  ylab("Residuals") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")
```

```{r echo=FALSE}
b <- allData %>%
  filter(allData$Saves != "N/A") %>%
  filter(Rk == "1") %>%
  mutate(year = as.numeric(year)) %>%
  select(Gls, year)

b %>%
  ggplot(aes(x = year, y = Gls)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)

lm1 <- lm(Gls~year, b)

b <- b %>%
  add_residuals(lm1) %>%
  add_predictions(lm1)

ggplot(b, aes(x=year, y =resid)) +
  geom_point() +
  xlab("x") +
  ylab("Residuals") +
  geom_hline(aes(yintercept=0), color="red", linetype = "dashed")
```

```{r echo=FALSE}
savesToGoals <- allData %>%
  filter(allData$Saves != "N/A") %>%
  filter(Rk == "1") %>%
  mutate(Squad =
    case_when(
      Squad == championNames$Squad[1] ~ Squad,
      Squad == championNames$Squad[2] ~ Squad,
      TRUE ~ "Other"
    )
  ) %>%
  group_by(year) %>%
  mutate(ratio = Saves/Gls) %>%
  select(ratio, year, Squad)

ggplot(savesToGoals, aes(x = year, y = ratio, color=Squad)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  ggtitle("Saves/Goals of Winners Over the Years") +
  ylab("Saves to Goals Ratio") +
  xlab("Year")

savesToGoals

s <- lm(Saves~Gls, allData)
summary(s)
```







