---
title: "STAT 240 Final Project"
output: html_document
---

## **Team Members**
Matthew Chiang, Siyu Huang, Vishnu Yaragadda

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
source("./scripts/viridis.R")
source("./scripts/ggprob.R")
```

```{r cache=TRUE, cache.path="cache/", include=FALSE}
# get all dates
file_list <- list.files(path="./data")

# load all data
allData <- data.frame()
for(i in 1:length(file_list)){
  
  # get file names
  league <- paste("./data/", file_list[i], "/LeagueTable.csv", sep="")
  standard <- paste("./data/", file_list[i], "/SquadStandardStats.csv", sep="")
  goalkeep <- paste("./data/", file_list[i], "/SquadGoalkeeping.csv", sep="")
  
  # load in each data table and merge it together
  LeagueTable <- read.csv(league)
  if( file.exists(standard) ){
    SquadStandardStats <- read.csv(standard)
    LeagueTable <- full_join(LeagueTable, SquadStandardStats, by="Squad")
  }
  if( file.exists(goalkeep) ){
    SquadGoalkeeping <- read.csv(goalkeep)
    LeagueTable <- full_join(LeagueTable, SquadGoalkeeping, by="Squad")
  }
  
  
  # add year column to the data
  LeagueTable <- select(LeagueTable, -contains(".y")) %>%
    mutate(year = substr(file_list[i], 6, 9))
  
  
  LeagueTable <- LeagueTable %>% 
    top_n(LeagueTable, 20)
  # combine the data into one big table
  if( i != 1){
    LeagueTable[setdiff(names(allData), names(LeagueTable))] <- NA
    allData[setdiff(names(LeagueTable), names(allData))] <- NA
  }
  allData <- rbind(allData, LeagueTable)
}
allData
```

## **Introduction**

Our team decided to conduct our analysis on the La Liga Football League as we all enjoy watching soccer and thought it would be interesting to try and make some sort of prediction on the winner of the current season. We decided that it would be best to analyze past seasons to predict the characteristics of the winner of the current season. 

To begin out approach we will look historicaly into the league to discover the probability that they win the current season. Then we will leverage a 95% confidence interval on a normal distribution of points and wins for the winners of the league. Lastly we will perform a linear regression on the winners of goals and saves vs years to exterpolate an estimation of how many saves and goals the 2020-2021 La Liga champion will have.

## **Background**

> Raw Data:

We found data for the La Liga Soccer League. Each season from 1988-2020, there are multiple sets of data ranging from team rankings, top scorers, standard squad statistics, team goalkeeping, team shooting, to statistics on individual players. We selected three sets of data:

* League Table: includes team rankings, total wins, loses, goals for, goals against, etc.
* Squad Standard Stats: includes general performance statistics of each team.
* Squad Goalkeeping: includes performance statistics for the goalkeeping of each team.

> Final Data: 

We had to collect the above data for the seasons between 1988-2020. We combined all of the data into one large dataframe by the team name and created a year column to keep track of which teams data cooresponds to which year.

> Data Citation:

"La Liga Seasons", Real Federación Española de Fútbol, 7th April 2020, https://fbref.com/en/comps/12/history/La-Liga-Seasons 
 
> Focus:
 
The overall aim of our analysis on this data will be to determine the characteristics of the winning team in the 2020-2021 La Liga Soccer League.


## **Analysis**


> To start our analysis, we wil look to see if the championship team's ratio of Goals Against to Saves is lower than non-championship teams.

To answer this question, we will plot Goals_Against/Saves over time for both Champion and non Champion team.
 
```{r echo=FALSE}
savesAndGoals <- allData %>%
  mutate(Rk =
    case_when(
      Rk == 1 ~ "Champion",
      TRUE ~ "Not Champion",
    ),
    ratio = GA.x/Saves
  ) %>%
  group_by(Rk) %>%
  select(Rk, year, ratio) %>%
  drop_na()
  
savesAndGoals %>%
  ggplot(aes(x=as.numeric(year), y=ratio, color=as.factor(Rk))) +
  geom_point(alpha = 0.5) +
  geom_smooth(method="lm", se = FALSE) +
  scale_color_manual(values = c("#e05a5e", "#58e0d5")) +
  labs(color = "Season Champion") +
  ggtitle("Goals Against/Saves vs Time", subtitle = "By Chamion and Non-Champion teams") +
  xlab("Year") +
  ylab("Goals Agaisnt over Saves (in a season)")
```
To check if the above graph is statistically significant we can perform a t test. If the p-value of the t test is less than our alpha (0.05) we can reject the null hypothesis.

```{r echo=FALSE}
champ <- savesAndGoals %>%
  filter(Rk == "Champion")

nonChamp <- savesAndGoals %>%
  filter(Rk == "Not Champion")

ttest = t.test(champ$ratio, nonChamp$ratio)
ttest
```
The p-value from our test was `r ttest$p.value`, which is less than 0.05 so we can reject the null hypothesis that the non-championship team's ratio equals the championship teams ratio. From the t test we can also say with 95% confidence that the championship teams ratio of Goals Against to Saves will be 0.061344 to 0.1301 below the ratio of non championship teams. This means that the number of Saves a team has in a season in more important because Saves is the denominator that is creating the smaller value, leading us to the conclusion that defensive plays a higher factor than offense.

> Another potential characteristic of the winner of a season is the proportion of goals by top scorer among total goals of the team. 

For each champion team of each season, they have a top scorer that scores more goals than other players in the team and we believe that this consists of a sizable proportion of total goals of the team comparing to other teams. 
```{R echo=FALSE}
top_scorer <- allData %>% 
  mutate(top_scorer_goals = as.numeric(str_trim(str_sub(Top.Team.Scorer, -2, -1)))) %>% 
  # mutate(t2 = str_extract(Top.Team.Scorer, "[\d+][?!.*\d]$")) %>% 
  mutate(prop = top_scorer_goals / GF) %>% 
  mutate(year = as.numeric(year)) %>% 
  select(Squad, top_scorer_goals, GF, prop, year)
# ggplot(top_scorer, aes(x=year, y=prop)) +
#   geom_point() +
#   geom_smooth(method = "lm", se = FALSE, color = "hotpink") +
#   ggtitle("The proportion of goals by top scorer in each seasons") +
#   labs(x="Season", y="The proportion of goals by top scorer")
```

```{R echo=FALSE}
top_scorer2 <- allData %>%
  mutate(Rk = case_when(Rk == 1 ~ "Champion", 
                        TRUE ~ "Not Champion",)) %>%
  mutate(top_scorer_goals = as.numeric(str_trim(str_sub(Top.Team.Scorer, -2, -1)))) %>% 
  mutate(prop = top_scorer_goals / GF) %>% 
  mutate(year = as.numeric(year)) %>% 
  select(Rk, top_scorer_goals, GF, prop, year)

not_champion_top_scorer <- top_scorer2 %>% 
  filter(Rk == "Not Champion") %>% 
  group_by(year) %>% 
  summarise(not_champion_prop = mean(prop))
champion_top_scorer <- top_scorer2 %>% 
  filter(Rk == "Champion") %>% 
  group_by(year) %>% 
  summarise(champion_prop = mean(prop)) %>% 
  inner_join(not_champion_top_scorer)
```

```{R echo=FALSE}
champion_top_scorer2 = gather(champion_top_scorer, event, total, champion_prop:not_champion_prop) %>% 
  mutate(`Season Champion` = event)
# champion_top_scorer2
ggplot(champion_top_scorer2, aes(x=`Season Champion`, y=total, fill=`Season Champion`)) +
  geom_boxplot() +
  ggtitle("Comparison between champions and other teams", subtitle = ":the proportion of goals by top scorer in each seasons") +
  labs(x="Champions or not", y="The proportion of goals by top scorer", color = "Season Champion") 
```


```{R echo=FALSE}
# lm_s = lm()
t.test(champion_top_scorer$champion_prop, champion_top_scorer$not_champion_prop, alternative = "greater")

# top_scorer2 %>% 
#   group_by(Rk) %>% 
#   summarise(sum_top_scorer = sum(top_scorer_goals),
#             sum_GF = sum(GF))
# prop.test(c(858, 8118), c(2794, 29612))
```

From the observation of the graph, we assume that the proportion of top scorer's goals in champions teams tend to be higher than those of non-champions teams. To verify our guess, we perform the hypothesis test where the null hypothesis is that the proportions are equal.  After doing the t-test, we found that the p-value is 0.01166 and it is statistically significant so that we can reject the null hypothesis. Therefore, The proportion of top scorer's goals of all goals in champions teams tend to be higher than those of non-champions teams.

<!-- > How about the trend of this proportion we just talked about? -->

```{R echo=FALSE}
# ggplot(top_scorer2, aes(x=year, y=prop, color=as.factor(Rk))) +
#   geom_point(alpha = 0.7) +
#   geom_smooth(method="lm", se = FALSE) +
#   scale_color_manual(values = c("darkgoldenrod", "blue")) +
#   ggtitle("The proportion of goals by top scorer in each seasons") +
#   labs(x="Season", y="The proportion of goals by top scorer", color = "Season Champion")
```

> One more potential characteristic of a winning team is the possession rate of that respective team. 

It's reasonable to assume that a winning team would have a higher possession rate as that would give the team more chances to score. So, we wanted to test that:

```{r, echo = FALSE}
poss <- allData %>%
  filter(Poss != "NA") %>%
  mutate(Rk = case_when(Rk <= 3 ~ "Top Teams", 
                        TRUE ~ "Rest of League",)) %>%
  group_by(Rk) %>%
  ggplot(aes(x = Poss, y = GF, color=as.factor(Rk))) +
  geom_point() +
  labs(x="Possession Rate", y="Goals Scored", color = "La Liga Teams") +
  ggtitle("Possession Rate vs Goals")

poss
  
```
The graph above supports our initial claim as the Top Teams have a much higher possession rate and thus end up with more goals scored. Now, we want to conduct a hypothesis test to reaffirm our claim that teams with high possession rates tend to do better. 

We can frame this as a hypothesis test described as follows:
$$
H_0: \mu_0 = \mu_1
$$
where $\mu_0$ is the mean possession rate for the rest of the league, and $\mu_1$ is the mean possession rate for the top teams.

$$
H_a: \mu_0 < \mu_1
$$

To carry out this test, we will consider a two-sample test with the following test statistic:
$$
T = \frac{(\bar{x}_0-\bar{x}_1) - (\mu_0-\mu_1)}{\sqrt{\frac{s_0^2}{n_0} + \frac{s_1^2}{n_1}}}
$$
where under the null $\mu_0 - \mu_1 = 0$, $n_i$ is the sample size for group $i=0,1$, and $s_i$ is the standard deviation of group $i$.

Since we are estimating the standard error, we will be using a t-distribution rather than a normal distribution.

```{r, echo = FALSE}
poss0 <- allData %>%
  filter(Poss != "NA") %>%
  filter(Rk > 3) %>%
  pull(Poss)

poss1 <- allData %>%
  filter(Poss != "NA") %>%
  filter(Rk <= 3) %>%
  pull(Poss)

mean0 <- mean(poss0)
sd0 <- sd(poss0)
n0 <- length(poss0)

mean1 <- mean(poss1)
sd1 <- sd(poss1)
n1 <- length(poss1)

## Assuming the null is true
se <- sqrt(sd0^2/n0+sd1^2/n1)
test_stat_poss <- (mean0-mean1)/se
test_stat_poss
```
```{r}
## Get degrees of freedom of t-distribution from t.test()
t.test(poss0, poss1, alternative ="less")

dof <- 19.6

pt(test_stat_poss, dof) # p-value = P(T_dof <= -9.76586)

```

Using a t-distribution with 19.6 degrees of freedom, we calculate a p-value on the order of $10^{-5}$ suggesting that we can reject the null hypothesis. This rejections means we have concluded that the mean possession rate for the top teams is greater than the mean possession rate for the rest of the league.

## Conclusion

One short coming of our project was that we had very limited amount of data. We only had about 20 years of data to work with which is not that much so our results might not be very accurate. Another short coming is that we tended to focus on only two aspects of the team in our linear regression. But, in reality, there might be a lot more variables that come into play. Instead of a linear regression, a multiple linear regression might suit this situation more, but we haven't learned that model yet.

When considering what makes a good soccer team, there are obviously more factors than just the number of goals and save they have. For example we could consider how a teams top player affects their performance within the league and factor that into our final analysis of the data. If we were to do this, we would find more data on specific player statistics, (i.e. Chances created, Shots On Goal, Possession Rate, etc.). We could also consider each teams overall offensive and defensive ratings and see how that has affected the winners of each season throughout history.